Chaotic mixing of tidal debris

### 2015-06-24 ms.tex
introduction take 2

As a satellite galaxy or globular cluster orbits within some larger system, mass is eroded due to the tidal forces of the host galaxy potential. If the orbit of such an object is regular, only mildly eccentric, and stays far from sharp density features (e.g., the disk), the disruption is a steady process and the mass is slowly stretched into long streams of matter known as stellar tidal streams. The phase-space density and therefore the observable properties of these tidal streams are sensitive to the internal progenitor properties, orbit of the progenitor, and mass distribution of the host galaxy. Many tidal streams are observed around the Milky Way, M31, and other nearby galaxies that span a large range of distances \citep[$\approx$10--100~kpc][]{cite many}. The morphological evolution of tidal debris depends on the spread of orbital properties (e.g., actions or frequencies) of the debris and the orbit of the progenitor system, both of which are also determined by the shape and radial profile of the gravitational potential of the host galaxy. By simultaneously modeling the observed phase-space density of stream stars along with the potential, it is hoped that we may infer the dark matter distribution around the Milky Way.

Tidal streams are generally modeled as 1D structures formed on regular orbits with simple internal progenitor dynamics and density evolution. Methods span a range of complexity from orbit-fitting, to \emph{Streakline} or particle-spray models, to action-space density modeling, to N-body simulations \citep[see, e.g., the introduction of][]{kuepper15}. All methods have been tested in some way on simulated observations of data and these tests typically demonstrate the recovery of parameters for analytic, static potential forms. Many of the methods have been shown to work in axisymmetric or otherwise simple potentials but may extend to more complex, generalized potentials (e.g., triaxial or multi-component potentials).

One example of stream modeling in a multi-component (static, analytic) potential was done by \citet{pearson15}, who tried to reproduce observations of the stellar stream density from the globular cluster Palomar 5 in a single oblate and single triaxial potential using \emph{Streakline} \citep{kuepper12} and N-body models. They use the observed SDSS number density of stream stars, a limited number of radial velocities for stream members, and the sky position, distance, radial velocity, and proper motion of the cluster itself to fit model streams to the data. In the oblate potential (a three-component bulge+disk+spherical halo potential), a thin model stream is easily found that reproduces the observed stellar density morphology of the stream. In the triaxial potential (the potential from \cite{law10}: a three component bulge+disk+triaxial halo fit to Sagittarius stream data), the model streams generically form large, two-dimensional ``fans'' of debris near the ends, and there are no physically reasonable progenitor orbits that reproduce the observed thinness of the stream given the observational constraints of the present-day position and velocity of the cluster.

The obvious difference between the two potentials considered by \citet{pearson15} is the extra symmetry of the triaxial potential. It is well known that the number of degrees of freedom of a potential plays a critical role in determining the orbit structure of the potential --- Hamiltonians with more than two degrees of freedom generically contain significant chaotic regions. \citet{pearson15} tested the stochasticity of the orbit of the progenitor that produced fanned debris by computing the Lyapunov exponent along this orbit but found that it is consistent with being regular over dynamically relevant timescales (many Hubble times). It has been shown previously that along some strongly chaotic orbits, tidal streams do form large, diffuse fans of debris \citep[e.g.,][]{fardal14}, however it is unknown how the resultant properties of the debris (e.g., density or length of the stream) depend on the degree of stochasticity. The result from \citet{pearson15} suggests that even weak chaos (as measured by the Lyapunov exponent) may affect the density evolution and therefore observability of tidal streams.

In this work, we study the effect of chaotic diffusion on tidal stream morphology. We choose a simple, cosmologically motivated model for a triaxial potential, analyze the degree of chaos as computed from single-orbit diagnostics for grids of constant-energy orbits, and compare these results to measures of the density evolution of finite-volume ensembles of orbits (meant to mimic tidal debris). We find that even when the chaotic timescale is predicted to be long for a given orbit, chaos may manifest over much shorter times in small orbit ensembles due to the nature of chaotic diffusion. This result supports a reevaluation of the importance of chaos in galactic halos and implies that the amount of chaos in a given potential may have significant consequences for the observability and survivability of thin, cold tidal streams.

>>>>

As a satellite galaxy or globular cluster orbits within some larger system, mass is eroded due to the tidal forces of the host galaxy potential. Along regular, mildly-eccentric orbits, mass is disrupted with small spreads in orbital properties (e.g., energy, angular momentum). Once the debris has evolved far enough from the progenitor system that the self-gravity of the progenitor can be ignored (usually a fast process relative to the orbital time), the stars evolve essentially as an ensemble of test particle orbits in the potential of the host system. The debris remains coherent as a tidal stream if the phase-mixing time-scale is long: a small ensemble of regular orbits reaches a fully phase-mixed state in a timescale $\approx\sigma_\Omega^{-1}$, where $\sigma_\Omega$ is the dispersion in fundamental frequencies of the ensemble. The ensemble spreads due to sheering from slight variations in their fundamental frequencies which, for tidal streams, preferentially occurs along one dimension \citep{merritt96, helmi99}.

The morphological (density) evolution of the tidal debris therefore depends only on the spread of orbital properties (e.g., actions or frequencies) of the debris and the orbit of the progenitor system, both of which are also determined by the shape and radial profile of the gravitational potential of the host galaxy. By modeling the observed phase-space density of stream stars along with the host galaxy potential it is hoped that we may infer the 3D mass distribution of the host. Many tidal streams are observed around the Milky Way, M31, and other nearby galaxies; the known streams span a large range of distances --- from $\approx10$ to $100~kpc$ --- and progenitor masses --- from $\approx$10$^3$ to $\approx$10$^8~\msun$ in stellar mass --- \citep[][]{cite many}. There has been extensive work on developing methods to use data from these streams to measure properties of Milky Way's dark matter halo. These methods span a range of complexity from orbit-fitting, to \emph{Streakline} or particle-spray models, to action-space density modeling, to N-body simulations \citep[see, e.g., the introduction of][]{kuepper15}. All methods have been tested in some way on simulated observations of data and these tests typically demonstrate the recovery of parameters for analytic, static potential forms. \todo{apw}{This is all just about the methods -- do I need sentences or a paragraph on results?}

%At present most of these methods have been shown to work in axisymmetric or otherwise simple potentials through tests with mock data.

One example of stream modeling in a multi-component (static, analytic) potential was done by \citet{pearson15}, who tried to reproduce observations of the stellar stream density from the globular cluster Palomar 5 in a single oblate and single triaxial potential using \emph{Streakline} \citep{kuepper12} and N-body models. They use the observed SDSS number density of stream stars, a limited number of radial velocities for stream members, and the sky position, distance, radial velocity, and proper motion of the cluster itself to fit model streams to the data. In the oblate potential (a three-component bulge+disk+spherical halo potential), a thin model stream is easily found that reproduces the observed stellar density morphology of the stream. In the triaxial potential (the potential from \cite{law10}: a three component bulge+disk+triaxial halo fit to Sagittarius stream data), the model streams generically form large, two-dimensional ``fans'' of debris near the ends, and there are no physically reasonable progenitor orbits that reproduce the observed thinness and curvature of the stream given the observational constraints of the present-day position and velocity of the cluster.

The obvious difference between the two potentials considered by \citet{pearson15} is the extra symmetry of the triaxial potential. It is well known that the number of degrees of freedom of a potential plays a critical role in determining the orbit structure of the potential --- Hamiltonians with more than two degrees of freedom generically contain significant chaotic regions. \citet{pearson15} tested the stochasticity of the orbit of the progenitor that produced `fanned' debris by computing the Lyapunov exponent along this orbit but found that it is consistent with being regular over dynamically relevant timescales (many Hubble times). It has been shown previously that along some strongly chaotic orbits, tidal streams do form large, diffuse `fans' of debris \citep[e.g.,][]{fardal14, ngan15}, however it is unknown how the resultant properties of the debris (e.g., density or length of the stream) depend on the degree of stochasticity. The result from \citet{pearson15} suggests that even weak chaos (as measured by the Lyapunov exponent) may affect the density evolution and therefore observability of tidal streams.

In this work, we study the effect of chaotic diffusion of individual orbits on the density evolution of tidal debris. We choose a simple, cosmologically motivated model for a triaxial potential, analyze the degree of chaos as computed from single-orbit diagnostics for grids of constant-energy orbits, and compare these results to measures of the density evolution of finite-volume ensembles of orbits (meant to mimic tidal debris). We find that even when the chaotic timescale is predicted to be long for a given orbit, chaos may manifest over much shorter times in small orbit ensembles due to the chaotic diffusion of the individual orbits. The diffusion of the fundamental frequencies of the orbits introduces an extra, time-dependent dispersion to the ensembles and therefore reduces the timescale over which mixing becomes relevant (e.g., $\sigma_\Omega$ can grow with time). This result supports a reevaluation of the importance of chaos in galactic halos and implies that the amount of chaos in a given potential may have significant consequences for the observability and survivability of thin, cold tidal streams.
